Merging merges, more or less

### 2015-04-08 2015mergingmerges/section/01-Introduction.tex
intro

Despite the years of work on array combinator fusion, scheduling dataflow networks and streaming languages, it seems one combinator has been cast aside as jetsam: the humble merge join.
Unlike the simple access patterns of append or zip, the merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.

@@

\TODO{Insert key example demonstrating some function involving merges that needs to be fused.}
\TODO{The intro, including the following bulleted list should fit all on the first page.}

We make the following contributions:
\begin{itemize}
\item a kind of deterministic finite automata, for expressing streaming computations with value-dependent rates (\S\ref{s:Machines});
\item some common combinators implemented with such machines (\S\ref{s:Machines:Combinators});
\item invariants required of such machines (\S\ref{s:Machines:Invariants});
\item and an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging}).
\end{itemize}

\subsection{Buffering}

Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
Consider the following program.
Suppose we have two sources of integers from disk, wish to filter each and pair them together, then perform some action for each pair:

\begin{code}
zip1 :: [Int] -> [Int] -> IO ()
zip1 xs ys
 = let xs'  = filter (>0) xs
       ys'  = filter (<0) ys
       zz   = zip xs' ys'
   in  mapM_ print        zz
\end{code}

This program can be executed with no buffering of either input or any intermediate arrays.
Now suppose wish to perform another action over the first source:

\begin{code}
zip2 :: [Int] -> [Int] -> IO ()
zip2 xs ys
 = let xs'  = filter (>0) xs
       ys'  = filter (<0) ys
       xs2  = map    (+1) xs
       zz   = zip xs' ys'
   in  do   mapM_ print   zz
            mapM_ print   xs2
\end{code}

If we were to execute this program either strictly or lazily, we must buffer the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ and performing the second action afterwards.
However, we could execute this with no buffering if the relative ordering of the two @mapM_@s were not important, and could be reordered and interspersed as the data is read.
We change the input lists to @Source@s and the @mapM_@s to @Perform@s, to denote that the IO actions are reorderable and to execute whenever data is available, and use @:&@ to join performs together.

\begin{code}
zip3 :: Source Int -> Source Int -> Perform
zip3 xs ys
 = let xs'  = filter (>0) xs
       ys'  = filter (<0) ys
       xs2  = map    (+1) xs
       zz   = zip xs' ys'
   in  Perform print      zz
   :&  Perform print      xs2
\end{code}

As a final example, suppose we removed the second data source and filtered the same source according to different predicates, then paired them together.
Can we still execute this with no buffer, and only iteration over @xs@?

\begin{code}
zip4 :: [Int] -> IO ()
zip4 xs
 = let xs'  = filter (>0) xs
       xs'' = filter (<0) xs
       zz   = zip xs' xs''
   in  mapM_ print        zz
\end{code}

Sadly this last program requires an unbounded buffer or two iterations over @xs@, as we do not know anything about the contents of @xs@.
Suppose that the entire list were made of a hundred positive numbers, followed by a hundred negative numbers: the first positive number would need to be kept around until the first negative number was reached, and likewise for the second and subsequent positive numbers -- so all one hundred positive numbers would need to be stored in memory until the negative numbers were reached.
Now consider if, instead of a hundred positive numbers, there were a thousand: now it is necessary to buffer the thousand numbers in memory instead.
The important part here is that one cannot compute, just by looking at the program, an adequate buffer size that works for all inputs.
We wish to disallow this program and others like it, where we cannot ensure execution with a bounded buffer.

>>>>

Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.

In addition, despite the years of work on array combinator fusion, scheduling dataflow networks and streaming languages, it seems one combinator has been cast aside as jetsam: the humble merge join.
Unlike the simple access patterns of append or zip, merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.

Merge joins are the heart of database joins, where two tables are ordered by the same key.
They are the ``conquer'' part of the divide and conquer merge sort.
They can even be used to implement segmented appends.

@@

For example, if we have two sets of values over time, one useful operation is a correlation: to find the intersection of the values at given times, and compute some new value based on the two values:
\begin{code}
correlate :: [(Date,Value)] -> [(Date,Value)]
          -> [(Date,Value)]
correlate c1 c2
 = let m1 = merge c1 c2
       g1 = group_by compute m1
   in  g1
\end{code}

The core of this algorithm is the @merge@ which intersperses the two inputs so equal dates are next to each other.
Now suppose we have a third set and wish to compute the correlation of the first and second, and the second and third.
Since we are already iterating over the second set, it would be ideal to compute this other correlation in the same iteration, fusing both correlations into a single loop.

\begin{code}
corr2     :: [(Date,Value)] -> [(Date,Value)]
          -> [(Date,Value)]
          ->([(Date,Value)],   [(Date,Value)])
corr2     c1 c2 c3
 = let m1 = merge c1 c2
       m2 = merge c2 c3
       g1 = group_by compute m1
       g2 = group_by compute m2
   in (g1, g2)
\end{code}

If this program were executed strictly, the first merge would be computed in total, reading @c1@ and @c2@, then the second merge would be computed, re-reading @c2@ from the start, and reading @c3@.
Laziness is no panacea either, as whether @c2@ may still be read if @g1@ were consumed in total, then @g2@ consumed.
Neither alone is sufficient to implement this as a single iteration.
Instead, the two merges are fused, and whenever either merge requires input from @c2@, we attempt to execute the other merge until it too attempts to read from @c2@.

We make the following contributions:
\begin{itemize}
\item a kind of deterministic finite automata, for expressing streaming computations with value-dependent rates (\S\ref{s:Machines});
\item some common combinators implemented with such machines (\S\ref{s:Machines:Combinators});
\item an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging});
\item invariants required of such machines (\S\ref{s:Machines:Invariants});
\item and proofs of correctness (\S\ref{s:Proofs});
\end{itemize}

\newpage
